Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2024-01-25 E-Hypergraphs-LICS24.tex
Some fixes

The translation from e-graphs to enriched string diagrams is illustrated informally below,
noting how the typing constraints of the $+$ constructor are satisfied in the image of the translation by discarding in each component all unnecessary inputs.
\[
    \tikzfig{categorical-semantics/egraph-translation-1}
\]
Examples of this translation are given in the second row of Figure \ref{fig:e-graph-example}. In particular, note that the translation of the \textit{cyclic} e-graph (e)
requires the use of a categorical \textit{trace}, generating a cycle. In this paper, we focus on \textit{acyclic} e-graphs and their corresponding categories; but the above example illustrates that the extension to the cyclic case seems to be routine. As it would involve needless additional presentational complexity we leave it as further work.

@@

We are interested not only in the free SMC over a set of generators, but also in \textit{(symmetric) monoidal theories} with extra equations such as, for instance, the rewrite rules (b)--(e) of Figure \ref{fig:e-graph-example}. These can be seen as equations between string diagrams, and thus rewrites of their hypergraph representations.  Because the generating equations can be applied in any context, we are lead to the notion of \textit{(hyper)graph rewriting}: given an equation $l=r$, we require some way to identify (a sub-hypergraph corresponding to) $l$ within a hypergraph and replace it with (a sub-hypergraph corresponding to) $r$.
The standard methodology to giving specifications of such graph rewriting, known as \textit{double pushout (DPO) rewriting} \cite{dpo, bonchi_string_2022} is still applicable.
However, these concepts must be now generalised from hypergraphs to \textit{e-hypergraphs}, hypergraphs with two additional relations denoting the hierarchical structure introduced by so-called \textit{e-boxes}: the generator for semilattice enrichment, and the separation of the components of each e-box.
Giving the appropriate definitions turns out to be a technical matter of some complexity, and constitutes the main body of the paper.

@@

    % \update{Can we check the proof? I haven't used all the conditions from monogamous-ness above.
    % Although, if the morphism $i+j \to L$ is mono, the complement is unique for not-monogamous hypergraphs as well.}

@@

    Consider the rule $\llbracket l \rrbracket = \llbracket r \rrbracket$ in Figure~\ref{fig:structural_rule_pushout} which corresponds to the top-left equation in Figure~\ref{fig:string-equations}.
    And recall the diagram assuming $\mathcal{L} = \llbracket l \rrbracket$, $\mathcal{R} = \llbracket r \rrbracket$, $\mathcal{G} = \llbracket g \rrbracket$ and $\mathcal{H} = \llbracket h \rrbracket$.
    \[
        \scalebox{0.8}{
        \tikzfig{combinatorial_semantics/DPOI_square}
        }
    \]
    We then further note that if $l$ is a subterm of $g$ then the vertices in the image of input and output external interfaces of $\llbracket l \rrbracket$ in $\llbracket g \rrbracket$ all share the same set of parents and are pairwise consistent if the set of parents is non-empty.
    This satisfies conditions (1)-(2) in Proposition~\ref{prop:complement_existence}.
    The no-dangling condition is satisfied because we consider only mda-cospans.
    Therefore, there exists a unique pushout complement $\mathcal{C}$ which is $\llbracket g \rrbracket$ with $\llbracket l \rrbracket$ removed.
    Then necessarily vertices in $\mathcal{C}$ in the image of $i+j$ share the same set of parents and are pairwise consistent as they do so in $\mathcal{G}$.
    The same is true for the image of $i + j$ in $\mathcal{R}$ and therefore the necessary pushout exists which is $\mathcal{C}$ with $\mathcal{R} = \llbracket r \rrbracket$ glued in and it is equal to $\llbracket h \rrbracket$.
    Hence $\llbracket g \rrbracket \Rrightarrow{} \llbracket h \rrbracket$.
    The rest follows by induction on the number of applied equations.
    Similar argument then applies to other rules in $\mathcal{S}$ as well.
\end{proof}

>>>>

The translation from e-graphs to enriched string diagrams is illustrated informally below,
noting how the typing constraints of the $+$ constructor are satisfied in the image of the translation by discarding in each component all unnecessary inputs.
\[
    \tikzfig{categorical-semantics/egraph-translation-1}
\]
Examples of this translation are given in the second row of Figure \ref{fig:e-graph-example}.
% In particular, note that the translation of the \textit{cyclic} e-graph (e) requires the use of a categorical \textit{trace}, generating a cycle. 
% In this paper, we focus on \textit{acyclic} e-graphs and their corresponding categories; but the above example illustrates that the extension to the cyclic case seems to be routine. As it would involve needless additional presentational complexity we leave it as further work.

@@

We are interested not only in the free SMC over a set of generators, but also in \textit{(symmetric) monoidal theories} with extra equations such as, for instance, the rewrite rules (b)--(d) of Figure \ref{fig:e-graph-example}. These can be seen as equations between string diagrams, and thus rewrites of their hypergraph representations.  Because the generating equations can be applied in any context, we are led to the notion of \textit{(hyper)graph rewriting}: given an equation $l=r$, we require some way to identify (a sub-hypergraph corresponding to) $l$ within a hypergraph and replace it with (a sub-hypergraph corresponding to) $r$.
The standard methodology to giving specifications of such graph rewriting, known as \textit{double pushout (DPO) rewriting} \cite{dpo, bonchi_string_2022} is still applicable.
However, these concepts must be now generalised from hypergraphs to \textit{e-hypergraphs}, hypergraphs with two additional relations denoting the hierarchical structure introduced by so-called \textit{e-boxes}: the generator for semilattice enrichment, and the separation of the components of each e-box.
Giving the appropriate definitions turns out to be a technical matter of some complexity, and constitutes the main body of the paper.

@@

Practically, the above proposition means that for any given match $m$ there is only one possible result of rewriting.

@@

    Consider the rule $\llbracket l \rrbracket = \llbracket r \rrbracket$ in Figure~\ref{fig:structural_rule_pushout} which corresponds to the top-left equation in Figure~\ref{fig:string-equations}.
    And recall the diagram assuming $\mathcal{L} = \llbracket l \rrbracket$, $\mathcal{R} = \llbracket r \rrbracket$, $\mathcal{G} = \llbracket g \rrbracket$ and $\mathcal{H} = \llbracket h \rrbracket$.
    \begin{figure}[h!]
    \[
        \scalebox{0.8}{
        \tikzfig{combinatorial_semantics/DPOI_square}
        }
    \]
    \end{figure}
    We then further note that if $l$ is a subterm of $g$ then the vertices in the image of input and output external interfaces of $\llbracket l \rrbracket$ in $\llbracket g \rrbracket$ all share the same set of parents and are pairwise consistent if the set of parents is non-empty.
    This satisfies conditions (1)-(2) in Proposition~\ref{prop:complement_existence}.
    The no-dangling condition is satisfied because we consider only mda-cospans.
    Therefore, there exists a unique pushout complement $\mathcal{C}$ which is $\llbracket g \rrbracket$ with $\llbracket l \rrbracket$ removed.
    Then necessarily vertices in $\mathcal{C}$ in the image of $i+j$ share the same set of parents and are pairwise consistent as they do so in $\mathcal{G}$.
    The same is true for the image of $i + j$ in $\mathcal{R}$ and therefore the necessary pushout exists which is $\mathcal{C}$ with $\mathcal{R} = \llbracket r \rrbracket$ glued in and it is equal to $\llbracket h \rrbracket$.
    Hence, $\llbracket g \rrbracket \Rrightarrow{} \llbracket h \rrbracket$.
    The rest follows by induction on the number of applied equations.
    Similar argument then applies to other rules in $\mathcal{S}$ as well.
\end{proof}

% \begin{proposition}[Proposition~\ref{prop:wnormal_form} restatement]
%     For each cospan $f$ in ${\WellTypedMdaEcospans}/{\mathcal{S}}$ there is a \textit{weak normal form} such that 
%     \[
%         f = f_1 + \ldots + f_n
%     \]
%      such that each $f_i$ contains no hierarchical edges,  and for all $i \neq j$ we have $f_i \neq f_j$.
% \end{proposition}
% \begin{proof}
%     We will define the depth of an edge $e$ as the number of parents it has, \textit{i.e.}, by $|[e)|$.
%     Then we will build the required weak normal form in steps and first will build a form $f = f_1 + \ldots + f_n$ such that every of $f_i$ is also of the form $g_1 + \ldots + g_k$ and so on and we will assume that $f$ has at least one hierarchical edge (otherwise is is already a weak normal form).
%     In order to achieve the latter we will apply rewrite rules that correspond to equations (1)-(4) from Figure~\ref{fig:string-equations}.
%     At step 1 we search for matches among the edges with depth 0 and then note that each such rewrite reduces the number of such edges until there is only one edge of depth 0, and it is necessarily a hierarchical one: because we have a finite number of edges of each depth, this process terminates.
%     At step $k$ we similarly search for matches among the edges with depth $k-1$ noting that this process terminates when we end up with $l$ groups of edges of depth $k-1$ such that edges within a group are pair-wise not consistent.
%     Because we do not increase the maximum depth in $f$ we will eventually reach a maximum value of $k$ and stop because there will be no hierarchical edges of depth $max(k)$ (otherwise their children would have depth $max(k) + 1$).

%     Then we will apply rules (5), from right to left, (6) and (8), from right to left.
%     Note that each of these rules decreases the number of hierarchical edges, so we apply them until there is only one hierarchical edge of depth 0, \textit{i.e.}, the top-level one.
%     This brings us to a weak normal form $f = f_1 + \ldots + f_n$ where each of $f_i$ does not contain any hierarchical edges and for $i \not = j$, $f_i \not = f_j$, the latter is guaranteed by rule (8).
% \end{proof}
